Asymmetric tail association in extreme climatic events affects population-extinction risk

### 2019-10-24 MainText.Rmd
response added

<!--***DAN: Dan has not edited this--> 
With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs 
could pose severe impact on the resistance or resilience of a population and/ or community level through
life-history traits of the species. Therefore predicting the impact of extreme events 
on population extinction risk is necessery. We here introduce the potential use of "tail association" 
property of two correlated ecological variables in modeling the extinction risk of 
an age-structured single patch model. We also extend this idea 
to density-dependent metapopulation models. For metapopulation models, 
we show extinction risk not only depends on the nature of tail association of 
spatially correlated noise but is also influenced by the inherent dynamics 
of the population-model. To illaborate, we demonstrate left tail dependent noise across spatially correlated patches 
cause higher extinction risk than the right tail dependent noise, when the model show undercompensatory dynamics.
But in overcompensatory regime, the trend is opposite. Our study paves the path for future research considering this tail association instead of simple correlation in population viability analysis.

@@

<!--***DAN: 
3) Please
add the figure you have mentioned in the text but that does not exist. Please run a spell
check. Please go and check every sentence throughout the manuscript for grammatical 
errors. I think you will find some 
and will be able to correct them if you focus on that - there are many, and by correcting them
you will not only make the paper better you will get better at writing grammatically correct 
sentences in English. 
-->

@@

<!--ECEs have a lot of impacts, it is important to understand their impacts-->
Over the past few decades, extreme climatic events (ECEs) such as
temperature or precipitation extremes, heat-waves, and extreme wildfires
have become more common or more extreme due to anthropogenic climate change 
[@ummenhofer2017extreme]. ECEs are well known to produce severe
impacts at the population, community and ecosystem levels [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@felton2017integrating], and thus the goal of improving our understanding of
the variety of ecological impacts of ECEs is an important applied goal for 
modern ecology [@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling]. Frequent extreme events 
can help alter the mean temperature or precipitation of an area, 
can be associated with changes through time in the overall variance of 
climate signals, and can directly contribute to changes in the symmetry or
skewness of distributions of climate signals through time 
[@hansen2012perception]. All such changes have the potential to influence 
populations, including their risk of extirpation from an area 
[@GarciaCarerras2013; @Vasseuretal2014].
For instance, @frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs on European shags 
via their effects on temporal variability in 
juvenile, immature and adult survival rates.
@mcdermott2017sensitivity investigated the sensitivity 
of $41$ UK butterfly species to four different types of ECEs. 
At the community level, research has showed that ECEs can reorder
the dominance heirarchy of species in an area, sometimes leading to 
selective local extictions, and can also produce effects that cascade 
across trophic levels [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
For example, @chiu2013impact showed that extreme floods affected 
the annual survival of a prey community, in turn influencing predators in 
aquatic as well as adjacent riparian habitats. 
@dreesen2014successive, in an experimental study, showed that 
successive ECEs with short time intervals between them
can weaken the resistance of plant communities.
<!--***DAN: Shyamolina, what, precisely, does it mean to "weaken the resistance of a
plant community"? Resistance to what? Please clarify.-->

@@

<!--concept for correlation in the extreme tail: tail association-->
Though earlier literature considered the influence of spatial synchrony 
in environmental variables on metapopulation extinction risk, to 
our knowledge the distinct influence of "asymmetric tail associations" between
environmental time series in different locations has hardly been considered. 
"Tail associations" are
associations between two random quantities in their extreme values. 
For instance, if two positively associated 
variables are more associated in their smaller values than in their larger values,
they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). 
If two positively associated variables are more associated in their larger 
values than in their smaller values, they are 
said to have stronger "right-tail" or "upper-tail" association 
(Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows that two variables can have a wide 
range of patterns of tail association, from extreme left- to extreme right-tail
association, with the same overall correlation. So tail association
is a distinct concept from correlation itself, and may have distinct ecological
ramifications [@Ghosh_copula]. Asymmetric tail associations have been documented 
between environmental variables, and between the same environmental variable 
measured in different places [@Serinaldi2008; @li2013; @She2018; @Goswami2018]. 
Asymmetric tail associations have also been documented between numerous ecological
variables and between environmental and ecological variables [@Ghosh_copula].

<!--ECEs probably produce tail associations-->
There are reasons to believe that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be stronger right-tail association between measurements of the climatic variable
made in different locations, through time.
For example, rain storms which deposit exceptionally large amounts of rain
may do so over large areas; but months with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater left-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because it that case it would be very good years for the component 
populations which would occur at the same time in many patches [@Ghosh_copulas].  
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modelling setup using a spatial version of a very simple non-density dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these cases. The present study extends tests of this idea to metapopulation
models incorporating density dependence.

Extinction risk is also very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002 ; @Caswell2006] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes very low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmtric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence extinction risk of a density dependendent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we use a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when overall correlation is kept fixed. 
For Q2, we consider spatial extensions of several simple and commonly
used population models. Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copulas). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are associated with
tail dependence in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Kendall 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases Kendall's $\tau$ was the same ($\tau \approx$ `r ktau_common`),
up to sample variation (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, can you please put the actual Kendall tau that pertained
for each dataset on each corresponding
panel? Use, say, 3-4 digits accuracy for display. These values should be very similar, 
but are expected to differ due to sampling variation. I would like to make that
clear.-->

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how do vital rates with asymmetric tail 
association affect the extinction risk of
populations in a single patch age-structured model,
we considered a simple stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$).

@@

We assumed the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear" with density dependence.
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail association between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail association had two broad 
categories: moderate and extreme. Each category had specific direction in tail association: 
either in left or in right tail of vital rates (as depicted in Fig. \ref{fig_intro_cop}). Vital 
rates ($f_a$ and $s_e$) were left tail dependent when unusual bad years ("catastrophes") 
suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

Moderate tail associatation was modeled by using the 
concept of "copula", a statistical tool to model 
dependencies. Appendix \ref{SM-intro_copula} summarized
a brief introduction of "copulas" and their relation to tail asscociations, symmetries 
with examples of available copula families 
from `R` packages (`copula` and `VineCopula`). We used bivariate Clayton 
copula to model moderate left-tail and Survival Clayton copula to model 
moderate right-tail association in the vital rates. Both copula had the same Kendall 
correlation ($\tau =$ `r ktau_common`) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail associated bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau =$ `r ktau_common`). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail association
with positive Kendall's correlation.

@@

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail association of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependent feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form: $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ dispersed equally to other $(n-1)$ patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and hence no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations 
between patches (to within sampling error)
for every simulation, but were moderately (or extremely) left- and 
right-tail associated between patches in
different simulations. For moderately left-tail associated 
noises we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($\tau =$ `r ktau_common`) between
any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail associated noises we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail association. We generated
extreme left- or right-tail associated noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
association in left and right-tail. <!--say how you set ext. threshold = Pe/10-->
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

The models stated in Table \ref{tab_model_summary} had one common 
feature: they showed undercompensatory
dynamics (monotonic approach towards stable steady state $P_e$) when
growth parameter $r$ remained below a critical 
level (say, $r_c$) and overcompensatory dynamics with oscillatory approach 
towards stable equilibrium $P_e$ when $r$ was beyond $r_c$. 
If $r$ became too high and exceeded $r_{bf}$ - the parameter value responsible for Hopf-bifurcation -
model equilibrium $P_e$ became unstable. Beyond Hopf-bifurcation point, model showed sustained
oscillation and eventually led to chaos with increasing $r$. Appendix \ref{SM-model_stability}
summarized the stability analysis of each model in their deterministic form.

@@

We plotted extinction risks along y-axis
for both left tail and right tail associated noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail associated noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risks at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risks from both noises overlapped
with each other. This could give us an idea of how extinction risks affected by the model parameter 
$r$ that controlled the underlying dynamics of the model, i.e., undercompensatory 
or overcompensatory.

@@

<!--Q1 : age structured results : linear and nonlinear model with asymmetric copula-->

>>>>

<!--***DAN: Dan has not edited this--> 
With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs could pose a severe impact on the resistance or resilience of a population and/ or community level through life-history traits of the species. Therefore predicting the impact of extreme events on population extinction risk is necessary. We here introduce the potential use of "tail association" property of two correlated ecological variables in modeling the extinction risk of an age-structured single patch model. We also extend this idea to density-dependent metapopulation models. For metapopulation models, we show extinction risk not only depends on the nature of the tail association of spatially correlated noise but is also influenced by the inherent dynamics of the population-model. To elaborate, we demonstrate left tail dependent noise across spatially correlated patches cause higher extinction risk than the right tail dependent noise when the model show undercompensatory dynamics. But in the overcompensatory regime, the trend is the opposite. Our study paves the path for future research considering this tail association instead of a simple correlation in population viability analysis.

@@

<!--***DAN: 
3) Please
add the figure you have mentioned in the text but that does not exist. Please run a spell
check. Please go and check every sentence throughout the manuscript for grammatical 
errors. I think you will find some 
and will be able to correct them if you focus on that - there are many, and by correcting them
you will not only make the paper better you will get better at writing grammatically correct 
sentences in English.

***Shya: 
Figure added, spell and grammar check done for the entire Maintext, but I think I should do it again just before submission.
-->

@@

<!--ECEs have a lot of impacts, it is important to understand their impacts-->
Over the past few decades, extreme climatic events (ECEs) such as
temperature or precipitation extremes, heat-waves, and extreme wildfires
have become more common or more extreme due to anthropogenic climate change 
[@ummenhofer2017extreme]. ECEs are well known to produce severe
impacts at the population, community and ecosystem levels [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@felton2017integrating], and thus the goal of improving our understanding of
the variety of ecological impacts of ECEs is an important applied goal for 
modern ecology [@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling]. Frequent extreme events 
can help alter the mean temperature or precipitation of an area, 
can be associated with changes through time in the overall variance of 
climate signals, and can directly contribute to changes in the symmetry or the 
skewness of distributions of climate signals through time 
[@hansen2012perception]. All such changes have the potential to influence 
populations, including their risk of extirpation from an area 
[@GarciaCarerras2013; @Vasseuretal2014].
For instance, @frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs on European shags 
via their effects on temporal variability in 
juvenile, immature and adult survival rates.
@mcdermott2017sensitivity investigated the sensitivity 
of $41$ UK butterfly species to four different types of ECEs. 
At the community level, research has shown that ECEs can reorder
the dominance hierarchy of species in an area, sometimes leading to 
selective local extinctions, and can also produce effects that cascade 
across trophic levels [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
For example, @chiu2013impact showed that extreme floods affected 
the annual survival of a prey community, in turn, influencing predators in 
aquatic as well as adjacent riparian habitats. 
@dreesen2014successive, in an experimental study, showed that 
successive ECEs with short time intervals between them
can weaken the resistance of plant communities.
<!--***DAN: Shyamolina, what, precisely, does it mean to "weaken the resistance of a
plant community"? Resistance to what? Please clarify.

***Shya: From @dreesen2014successive, authors concluded 
"Our results show that when experimental temperate herbaceous assemblages were exposed to an extreme climatic
event, their resistance to a subsequent event, in terms of leaf
survival, is reduced only when the recovery period is relatively short (14 days for the assemblages studied here) and
not when it is longer."-->

@@

<!--concept for correlation in the extreme tail: tail association-->
Though earlier literature considered the influence of spatial synchrony 
in environmental variables on metapopulation extinction risk, to 
our knowledge the distinct influence of "asymmetric tail associations" between
environmental time series in different locations has hardly been considered. 
"Tail associations" are
associations between two random quantities in their extreme values. 
For instance, if two positively associated 
variables are more associated with their smaller values than in their larger values,
they are said to have stronger
"left-tail" or "lower-tail" association (Figure \ref{fig_intro_cop}A, B). 
If two positively associated variables are more associated in their larger 
values than in their smaller values, they are 
said to have stronger "right-tail" or "upper-tail" association 
(Figure \ref{fig_intro_cop}C, D). 
Figure \ref{fig_intro_cop} shows that two variables can have a wide 
range of patterns of tail association, from extreme left- to extreme right-tail
association, with the same overall correlation. So the tail association
is a distinct concept from correlation itself and may have distinct ecological
ramifications [@Ghosh_copula]. Asymmetric tail associations have been documented 
between environmental variables and between the same environmental variable 
measured in different places [@Serinaldi2008; @li2013; @She2018; @Goswami2018]. 
Asymmetric tail associations have also been documented between numerous ecological
variables and between environmental and ecological variables [@Ghosh_copula].

<!--ECEs probably produce tail associations-->
There are reasons to believe that ECEs may produce asymmetric tail 
associations between climatic variables
measured in different locations. For instance, if a single ECE is associated
with spatially widespread, large values 
of some climatic variable, but small values of the
variable are associated with local weather rather than ECEs, the result could
be a stronger right-tail association between measurements of the climatic variable
made in different locations, through time.
For example, rainstorms which deposit exceptionally large amounts of rain
may do so over large areas, but months with less rain may experience
scattered showers only, which may be much less correlated across locales.

@@

There are also reasons to hypothesize that asymmetry of tail association in 
environmental variables may influence extinction risks for metapopulations.
For instance, assuming for the sake of a simple example that low values of an environmental
variable are "bad" for populations of a given species and high values are "good",
greater left-tail associations between measurements of the variable in different
habitat patches may cause higher metapopulation extinction risk because then
very bad years for component populations would occur at the same time in 
many patches, reducing the potential for rescue effects. Greater left-tail associations
would probably not have the same effect, even if overall correlations between locations
were the same, because it that case it would be very good years for the component 
populations that would occur at the same time in many patches [@Ghosh_copula]. 
@Ghosh_copula substantiated this intuition by showing that it held true for a 
modeling setup using a spatial version of a very simple non-density dependent
population growth model (the Lewontin-Cohen model). In essence, if environmental
"catastrophes" (i.e., extremely bad years for a population) are widely 
spatially synchronized it should create much greater metapopulation extinction 
risk than if "bonanzas" (i.e., extremely good years) are widely synchronized, even
if overall (non-tail-specific) levels of environmental synchrony are the same
in both these cases. The present study extends tests of this idea to metapopulation
models incorporating density dependence.

Extinction risk is also very commonly estimated for age-structured models 
such as stochastic matrix models [@MorrisDoak2002; @Caswell2006] and 
integral projection models
[@Merowetal2014]. For such models, extinction risk depends, in part, on the long-term
population growth rate, a population parameter which in turn 
depends on the means, variances and covariances of population vital rates such 
as the survival and fecundity rates of age classes. Extinction risks may,
*a priori*, also depend on possible asymmetries of tail association between
population vital rates. For instance, it seems reasonable to imagine
a case in which both adult fecundity and juvenile survival rates 
depend on an environmental variable such as spring rainfall in similar ways, but 
only when that variable takes very low values. When rainfall takes
typical or high values, fecundity and survival may be limited instead 
by different factors. This could produce left-tail dependence between these 
vital rates, possibly resulting in different extinction risk for the model
compared to a case in which vital rates were symmetrically associated in 
their tails but otherwise statistically similar.

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we use a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we consider spatial extensions of several simple and commonly
used population models. Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are associated with
tail dependence in environmental variables.

@@

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Kendall 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Kendall's $\tau$ was the same ($\tau \approx$ `r ktau_common`),
up to sample variation (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, can you please put the actual Kendall tau that pertained
for each dataset on each corresponding
panel? Use, say, 3-4 digits accuracy for display. These values should be very similar, 
but are expected to differ due to sampling variation. I would like to make that
clear.
***Shya: correlation added (rounded upto 4 digit) in each plot.-->

@@

<!--stochastic matrix model : age-structured-->
To investigate (Q1) from \nameref{Introduction} that how do vital rates with an asymmetric tail 
association affect the extinction risk of
populations in a single patch age-structured model,
we considered a simple stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$).

@@

We assumed the population was semelparous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear" with density dependence.
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail association between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail association had two broad 
categories: moderate and extreme. Each category had specific direction in tail association: 
either in left or in right tail of vital rates (as depicted in Fig. \ref{fig_intro_cop}). Vital 
rates ($f_a$ and $s_e$) were left tail dependent when unusual bad years ("catastrophes") 
suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

The moderate tail association was modeled by using the 
concept of "copula", a statistical tool to model 
dependencies. Appendix \ref{SM-intro_copula} summarized
a brief introduction of "copulas" and their relation to tail associations, symmetries 
with examples of available copula families 
from `R` packages (`copula` and `VineCopula`). We used bivariate Clayton 
copula to model moderate left-tail and Survival Clayton copula to model 
moderate right-tail association in the vital rates. Both copulae had the same Kendall 
correlation ($\tau =$ `r ktau_common`) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail associated bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau =$ `r ktau_common`). Entries within each column of
matrix $M$, then became independent of each other but 
entries between columns had a specific tail association
with positive Kendall's correlation.

@@

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how the nature of the tail association of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependent feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two: Varley model (this is not a continuous one) 
and the Malthus-Condorcet-Mill model (terms containing logarithm of the extinct population does not make sense). 
All models assumed a generalized form: $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. An n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ dispersed equally to other $(n-1)$ patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and hence no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations 
between patches (to within sampling error)
for every simulation, but were moderately (or extremely) left- and 
right-tail associated between patches in
different simulations. For moderately left-tail associated 
noises we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($\tau =$ `r ktau_common`) between
any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail associated noises we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail association. We generated
extreme left- or right-tail associated noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
association in left and right-tail. <!--say how you set ext. threshold = Pe/10-->
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed the population went extinct after each time step if it was below a 
certain threshold (set to 10\% of the initial population $P_e$ for individual model).

The models stated in Table \ref{tab_model_summary} had one common 
feature: they showed undercompensatory
dynamics (monotonic approach towards stable steady state $P_e$) when
growth parameter $r$ remained below a critical 
level (say, $r_c$) and overcompensatory dynamics with oscillatory approach 
towards stable equilibrium $P_e$ when $r$ was beyond $r_c$. 
If $r$ became too high and exceeded $r_{bf}$ - the parameter value responsible for Hopf-bifurcation -
model equilibrium $P_e$ became unstable. Beyond the Hopf-bifurcation point, the model showed sustained
oscillation and eventually led to chaos with increasing $r$. Appendix \ref{SM-model_stability}
summarized the stability analysis of each model in their deterministic form.

@@

We plotted extinction risks along the y-axis
for both left tail and right tail associated noises 
against the varying range of $r$ along x-axis when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal to avoid any rescue effect among patches. 
For each model with extreme (or moderate) left- and right-tail associated noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risks at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risks from both noises overlapped
with each other. This could give us an idea of how extinction risks affected by the model parameter 
$r$ that controlled the underlying dynamics of the model, i.e., undercompensatory 
or overcompensatory.

@@

<!--Q1 : age-structured results : linear and nonlinear model with asymmetric copula-->
